Nitric-oxide-feedback to ciliary photoreceptor cells orchestrates UV avoidance

### 2022-09-30 text/Manuscript.Rmd
Put supplement

The ability to perceive changes in the environment and adapt behaviour accordingly is essential for the survival of all living organisms. Avoidance from UV radiation, which damages DNA essential for survival, is proposed to be one of the behaviours that existed before the abundant development of Earth's oxygen to block UV radiation in the atmosphere (Pittendrigh). In the ocean, almost all zooplankton have a pronounced rhythmic behaviour called Diel vertical migration (DVM), which consists of upwards towards the surface at dusk and descend back to deeper water before dawn. Avoidance behaviour to escape harmful UV radiation during the day is closely linked to the control of vertical migration in the ocean. In metamorphosis, a life history strategy widely found in metazoans, downward swimming towards the seafloor is known to be associated with larval settlement and the initiation of metamorphosis, a dramatic body change from the larval stage, and UV-avoidance behaviour allows dispersal and the acquisition of different ecological niches. However, little is known about the molecular neural circuit mechanisms in UV avoidance behaviour in many animals, including zooplankton. In neural circuits in general light response, the activation of various receptors by excitatory and inhibitory neurotransmitters such as glutamate, GABA and dopamine is known in various species. In addition to this, it has long been known that in nitric oxide (NO), as a neuromodulatory molecule, it is also important as a modulator of visual activity. NO is a small free-radical molecule with diffusible and changeable characteristics, and thus works as one of the neural signalling molecules. The function of NO in the neural circuitry of vision has been well studied in vertebrates. In the mammalian retina, NO is an essential signalling molecule in all retinal cells. NO produced by nitric oxide synthase (NOS) in amacrine cells, one of the retinal cells, activates soluble guanylate cyclase (sGC) expressed in cone bipolar cells, which produces cGMP and regulates downstream signalling. We therefore hypothesise that NO-mediated neural signalling may be involved in the photoreceptor circuits that control UV avoidance behaviour. To test this hypothesis we used the annelid *Platynereis dumerilii*, which conserved vertebrate-type ciliary photoreceptors (cPRCs) [@Arendt2004]. The cPRC expresses c-opsin1, which absorbs UV radiation, is involved in melatonin production and regulates circadian swimming behaviour [@Tosches2014; @tsukamoto2017]. Furthermore, we previously reported abnormal UV avoidance behaviour in c-opsin1 knockout larvae [@Veraszto2018]. The cPRCs express the UV-absorbing c-opsin1 and mediate UV avoidance in the larvae [@Veraszto2018]. In adult worms, UV light and c-opsin1 modulate locomotion and maturation [@Veedin2021]. These suggest that Platynereis larvae are a highly suitable model for elucidating the neural circuit mechanisms of NO-induced UV avoidance behaviour, and the accessibility of genome editing, behavioural assays, calcium imaging and whole-body connectomics could be very useful for study at the molecular level[@ozpolat2021].

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data. Phylogenetic analysis of NOS proteins indicate that the *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences (Fig. 1---figure supplement 1). To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis. In two-day-old larvae, we detected *NOS* expression in two cells in the apical organ region (Fig. 1A) <!---# 3d HCR, Kei will put them -->. The same two cells cells were labelled with a reporter construct driving palmitoylated-tdTomato by a 12 kbp upstream genomic region of the *NOS* gene. This reporter also revealed the axonal projections of the NOS-expressing neurons. In three-day-old larvae... <!---# need 3d HCR , Kei will put them -->

The position and morphology of the four *NOS*-expressing cells allowed us to identify the same cells in the three-day-old whole-body volume EM data [@williams2017; @veraszto2020] as four interneurons (INNOS)(Figure 1D-E). The INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets (Fig. 1F).

![**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** (A) SEM image of a 72 hpf Platynereis larva.(B-D) Expression analysis of the NOS gene (magenta) using in situ HCR. Co-stained image using acetylated α-tubulin antibody (ac α-tub: green). Anterior view of the larva at 54 and 72 hpf. (E) Immunostaining with anti-HA antibody against HA tagged palmitoylated-tdTomato expressed under the upstream of the start site of NOS (magenta). Co-staining with acetylated α-tubulin antibody as marker for cilia and axonal scaffold (ac α-tub: green). INNOSs with the entire cell body and axon labelled are shown with white arrowheads and with only the cell body labelled are shown with black arrowheads. Views of larvae at 72 hpf from the anterior side. (F and G) Five types of neurons (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed in the whole-body transmission electron microscopy (ssTEM) dataset of a 72 hpf larva. The INNOS on the ventral side labelled with (E) are shown in purple. Axons and dendrites appear as lines and cell body positions are represented by spheres. (H) Wiring diagram of the cPRC circuit. <!---# This graph widget does not look very good. We need to edit this a bit more. --> Hexagons represent neuronal groups (numbers indicated in square brackets), arrows represent synaptic connections (numbers indicated in number of synapses); NOS and RGW indicate INNOS and INRGW-dc1 respectively. Ciliomotor neuron shows two serotonergic neurons (Ser-h1).](../figures/Fig1.png){width="100%"}

**Nitric oxide signalling mediates UV-avoidance behaviour** The specific expression of NOS in interneurons directly postsynaptic to the cPRCs suggests that NO signalling is involved in the UV response. To test this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method. We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to an early stop codon and thus likely representing null alleles (Figure 3A). We could establish a homozygous line for both mutations indicating that NOS is not an essential gene. To quantify UV avoidance, we recorded the trajectories of freely swimming larvae in vertical columns, illuminated laterally from two opposite sides with xy nm violet light. As previously shown, wild-type larvae swim downward within 30 sec following UV/violet light stimulation. In contrast, homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response (Figure 3B). This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae and reveals a requirement for NOS in UV-avoidance behaviour.

![**Figure 2. Larvae with inhibited NOS have shorter ciliary arrest times and a delayed response to UV-avoidance.** (A) Top: The domain structure of NOS. Bottom: Close-up region showing the genomic locus of NOS gene and the wild-type sequence (WT) targeted by CRISPR/Cas9. The generated mutants (NOSΔ11, NOSΔ23) are also shown. Pink indicates target sites. Yellow shows PAM sequences, red shows stop codons. (B and C) Vertical swimming in wild type and mutant larvae at 2-day-old (B) and 3-day-old (C) stimulated with blue (480 nm) and UV (395 nm) light from above. the data are shown in 30 s bins.](../figures/Fig2.png){alt="Figure 2. Larvae with inhibited NOS have shorter ciliary arrest times and a delayed response to UV-avoidance. (A) Percentage of time in which cilia are closed in DMSO-treated larvae (control) vs. larvae exposed to NOS inhibitor. (B) Top: The domain structure of NOS. Bottom: Close-up region showing the genomic locus of NOS gene and the wild-type sequence (WT) targeted by CRISPR/Cas9. The generated mutants (NOSΔ11, NOSΔ23) are also shown. Pink indicates target sites. Yellow shows PAM sequences, red shows stop codons. (C and D) Vertical swimming in wild type and mutant larvae at 2-day-old (C) and 3-day-old (D) stimulated with blue (480 nm) and UV (395 nm) light from above. the data are shown in 30 s bins." width="100%"}

@@

NO generally acts via cyclic guanosine monophosphate (cGMP) as a second messenger. In this signalling pathway, NO binds to the heme group of soluble guanylate cyclase (sGC), a member of the guanylate cyclase family with a CYC domain (PF00211), to increase cGMP. However, no NO targets were found in cells that highly express sGC within the cPRC circuit. Recently, Moroz and others reported that GC with NIT (nitrite/nitrate sensing) domains (PF08376) (NIT-GC), identified as a sensor for nitrate and nitrite, the oxides of NO, is relatively widely conserved in metazoans (Moroz et al., 2020). To search for NIT-GCs in Platynereis, genes with the CYC domains were searched using the transcriptome and genome sequences of 45 metazoan species and choanoflagellate. By sequence-similarity searches, multiple sequence alignments and cluster analyses, a cluster of membrane-bound (guanylate cyclase-coupled receptor) type was found and a phylogenetic tree was constructed with soluble type (sGC) as an outgroup. As in the phylogenetic analysis of Moroz et al, we found a group of GCs with NIT domains that included many metazoans such as placozoa, cnidaria, ecdysozoa, echinoderms and lophotrochozoa, and in Platynereis 15 NIT-GCs were found (Fig. 3A). To find the candidate NIT-GCs involved in the cPRC circuit among the 15 NIT-GCs, we used previously published spatially mapped single-cell transcriptome data [@Achim2015][@williams2017] and Two NIT-GCs were found to be highly expressed in the cPRC, and we analysed their expression using in situ HCR as NIT-GC1 and NIT-GC2. As a result, strong signals for both were observed specifically in the vicinity of the cPRCs in 2-day-old larvae (Fig. 3B). Both c-opsin1 and MLD pNP (pedal peptide neuropeptide precursor 2), already identified as cPRC markers, and NIT-GC1 and NIT-GC2 were found to be co-expressed. They were co-expressed c-opsin1 and MLD pNP (pedal peptide neuropeptide precursor 2), already identified as cPRC markers (Fig. 3C) (Arendt et al., 2004; Williams et al., 2017). To analyse localisations at the protein level, immunostaining was performed using polyclonal antibodies against NIT-GC1 and NIT-GC2. Interestingly, NIT-GC1 was found to localise to the neuronal axon and NIT-GC2 to the ciliary region of the cPRC (Fig. 3D and E). Furthermore, co-immunostaining with NOS antibodies revealed that NOS and NIT-GC1 were localised in close proximity in the neurosecretory plexus (Fig. 3F). These results suggest that NIT-GC1 may function as a major target for NO.

@@

NIT-GC1 has a highly conserved cyclase domain and activation of NIT-GC1 is expected to produce cGMP. Recently, a synthetic protein called Green cGull was developed as a cGMP indicator that can be used in cultured cell systems (Matsuda et al., 2017). We co-expressed Green cGull and NIT-GC1 in cultured cells COS-7 (monkey kidney) and tested the intracellular dynamics of cGMP in the activation of NIT-GC1 (Fig. 4A). First, S-nitroso-N-acetyl-DL-penicillamine (SNAP; NO donor), 8-Br-cGMP (8-Bromoguanosine 3',5'-cyclic monophosphate; membrane-permeable cGMP analogue, positive control) and DMSO (negative control) were applied to COS-7 cells expressing Green cGull only. 8-Br-cGMP increased fluorescence intensity by more than sixfold, while little changes were detected for SNAP and DMSO (Fig. S4A). Next, the two kind of genes, Green cGull and NIT-GC1, were co-expressed using the 2A self-cleaving peptide and the same solutions were applied. As a result, a clear increase in cGMP fluorescence intensity was detected in SNAP compared to DMSO or expression of Green cGull only, although not as high as the increase in 8-Br-cGMP (Fig. 4B, C). Finally, expression of a mutant form of NIT-GC1 with deletion of the NIT domain significantly suppressed the increase in cGMP (Note, we need re-experiment!, Fig. 4D). These results indicate that NIT-GC1 is activated by NO to produce cGMP. They also show that the NIT domain is required for the activation of NIT-GC1.

>>>>

The ability to perceive changes in the environment and adapt behaviour accordingly is essential for the survival of all living organisms. Avoidance from UV radiation, which damages DNA essential for survival, is proposed to be one of the behaviours that existed before the abundant development of Earth's oxygen to block UV radiation in the atmosphere (Pittendrigh). In the ocean, almost all zooplankton have a pronounced rhythmic behaviour called Diel vertical migration (DVM), which consists of upwards towards the surface at dusk and descend back to deeper water before dawn. Avoidance behaviour to escape harmful UV radiation during the day is closely linked to the control of vertical migration in the ocean. However, little is known about the molecular neural circuit mechanisms in UV avoidance behaviour in many animals, including zooplankton. In neural circuits in general light response, the activation of various receptors by excitatory and inhibitory neurotransmitters such as glutamate, GABA and dopamine is known in various species. In addition to this, it has long been known that in nitric oxide (NO), as a neuromodulatory molecule, it is also important as a modulator of visual activity. NO is a small free-radical molecule with diffusible and changeable characteristics, and thus works as one of the neural signalling molecules. The function of NO in the neural circuitry of vision has been well studied in vertebrates. In the mammalian retina, NO is an essential signalling molecule in all retinal cells. NO produced by nitric oxide synthase (NOS) in amacrine cells, one of the retinal cells, activates soluble guanylate cyclase (sGC) expressed in cone bipolar cells, which produces cGMP and regulates downstream signalling. We therefore hypothesise that NO-mediated neural signalling may be involved in the photoreceptor circuits that control UV avoidance behaviour. To test this hypothesis we used the annelid *Platynereis dumerilii*, which conserved vertebrate-type ciliary photoreceptors (cPRCs) [@Arendt2004]. The cPRC expresses c-opsin1, which absorbs UV radiation, is involved in melatonin production and regulates circadian swimming behaviour [@Tosches2014; @tsukamoto2017]. Furthermore, we previously reported abnormal UV avoidance behaviour in c-opsin1 knockout larvae [@Veraszto2018]. The cPRCs express the UV-absorbing c-opsin1 and mediate UV avoidance in the larvae [@Veraszto2018]. In adult worms, UV light and c-opsin1 modulate locomotion and maturation [@Veedin2021]. These suggest that Platynereis larvae are a highly suitable model for elucidating the neural circuit mechanisms of NO-induced UV avoidance behaviour, and the accessibility of genome editing, behavioural assays, calcium imaging and whole-body connectomics could be very useful for study at the molecular level[@ozpolat2021].

@@

We identified a single *nitric oxide synthase* (*NOS*) gene in the *Platynereis dumerilii* genome and transcriptome data. Phylogenetic analysis of NOS proteins indicate that the *Platynereis* NOS belongs to an orthology group of metazoan NOS sequences (Figure 1---figure supplement 1). To characterise the expression pattern of *NOS* we used in situ hybridization chain reaction (HCR) and transient transgenesis. In 2 and 3-day-old larvae, we detected *NOS* expression in two or four cells in the apical organ region (Figure 1A-D). The same four cells were labelled with a reporter construct driving palmitoylated-tdTomato by a 12 kbp upstream genomic region of the *NOS* gene in 3-day-old larvae(Figure 1E). This reporter also revealed the axonal projections of the NOS-expressing neurons. The position and morphology of the four *NOS*-expressing cells allowed us to identify the same cells in the 3-day-old whole-body volume EM data [@williams2017; @veraszto2020] as four interneurons (INNOS)(Figure 1F and G). The INNOS cells are postsynaptic to the UV-sensory cPRCs and presynaptic to the INRGW interneurons, which are also cPRC targets (Figure 1H).

![**Figure 1. Identification of NOS-expressing interneurons (INNOS) within the cPRC circuit.** (A) SEM image of a 3-day-old Platynereis larva.(B-D) Expression analysis of the NOS gene (magenta) using in situ HCR. (D) Co-stained image using acetylated α-tubulin antibody (ac α-tub: green). (B) Dorsal view of the larva at 3-day-old. (C and D) Anterior view of the larva at 2 (D) and 3-day-old (C). (E) Immunostaining with anti-HA antibody against HA tagged palmitoylated-tdTomato expressed under the upstream of the start site of NOS (magenta). Co-staining with acetylated α-tubulin antibody as marker for cilia and axonal scaffold (ac α-tub: green). INNOSs with the entire cell body and axon labelled are shown with white arrowheads and with only the cell body labelled are shown with black arrowheads. Views of larvae at 3-day-old from the anterior side. (F and G) Five types of neurons (cPRC, INNOS, INRGWa, Ser-h1 and MC) in the cPRC circuit reconstructed in the whole-body transmission electron microscopy (ssTEM) dataset of a 3-day-old larva. The INNOS on the ventral side labelled with (E) are shown in purple. Axons and dendrites appear as lines and cell body positions are represented by spheres. (H) Wiring diagram of the cPRC circuit. <!---# This graph widget does not look very good. We need to edit this a bit more. --> ](../figures/Fig1.png){width="100%"}

**Nitric oxide signalling mediates UV-avoidance behaviour** The specific expression of NOS in interneurons directly postsynaptic to the cPRCs suggests that NO signalling is involved in the UV response. To test this, we generated two *Platynereis* *NOS* knockout lines with the CRISPR/Cas9 method. We recovered two deletions (NOSΔ11/Δ11 and Δ23/Δ23), both leading to an early stop codon and thus likely representing null alleles (Figure 2A). We could establish a homozygous line for both mutations indicating that NOS is not an essential gene. To quantify UV avoidance, we recorded the trajectories of freely swimming larvae in vertical columns, illuminated laterally from two opposite sides with 395 nm violet light. As previously shown, wild-type larvae swim downward within 30 sec following UV/violet light stimulation [@veraszto2018]. In contrast, homozygous *NOS*-mutant larvae showed a strongly diminished UV-avoidance response (Figure 2B and C). This phenotype is similar to the defective UV-avoidance of *c-opsin1* mutant larvae and reveals a requirement for NOS in UV-avoidance behaviour [@veraszto2018].

To analyse the behaviour of larvae before and after UV stimulation in more detail, the temporal changes in some parameters were compared using tracking of individual larvae of the WT and mutant (Figure 2D). The results showed that in the WT there was a rapid and strong increase in downward swimming approximately 15 s after UV stimulation, whereas this did not occur in the mutant (Figure 2E). Interestingly, there was already a difference in swimming speed between the WT and the mutant before UV stimulation (in the dark) (Figure 2F). After UV stimulation, the swimming speed of the WT increased and reached a maximum after about 15 s. In contrast, the swimming speed of the mutant did not increase after about 15 s. In contrast, the mutants showed a slight increase in speed after UV stimulation, but swimming speed was not maintained and returned to the speed before UV stimulation. To investigate the role of NO on UV avoidance behaviour, receptors for NO were explored.

![**Figure 2. Larvae with inhibited NOS have shorter ciliary arrest times and a delayed response to UV-avoidance.** (A) Top: The domain structure of NOS. Bottom: Close-up region showing the genomic locus of NOS gene and the wild-type sequence (WT) targeted by CRISPR/Cas9. The generated mutants (NOSΔ11, NOSΔ23) are also shown. Pink indicates target sites. Yellow shows PAM sequences, red shows stop codons. (B and C) Vertical swimming in wild type and mutant larvae at 2-day-old (B) and 3-day-old (C) stimulated with blue (488 nm) and UV (395 nm) light from above. The data are shown in 30 s bins. (D) Overlaid trajectories for WT (left, n=20) and NOS mutant (right, n=31) 3-day-old larvae. Grey shows the trajectories before UV stimulation and purple shows the trajectories after UV stimulation. (E) The temporal changes in the vertical position of the WT and mutant larvae before and after UV stimulation are shown. The starting points of each larval trajectory are set to 0. (F) The temporal changes in the distance traveled of the WT and mutant larvae before and after UV stimulation are shown. (E and F) After UV stimulation is indicated by purple squares.](../figures/Fig2.png){width="100%"}

@@

NO generally acts via cyclic guanosine monophosphate (cGMP) as a second messenger. In this signalling pathway, NO binds to the heme group of soluble guanylate cyclase (sGC), a member of the guanylate cyclase family with a CYC domain (PF00211), to increase cGMP. However, no NO targets were found in cells that highly express sGC within the cPRC circuit. Recently, Moroz and others reported that GC with NIT (nitrite/nitrate sensing) domains (PF08376) (NIT-GC), identified as a sensor for nitrate and nitrite, the oxides of NO, is relatively widely conserved in metazoans [@Moroz2020]. To search for NIT-GCs in Platynereis, genes with the CYC domains were searched using the transcriptome and genome sequences of 45 metazoan species and choanoflagellate. By sequence-similarity searches, multiple sequence alignments and cluster analyses, a cluster of membrane-bound (guanylate cyclase-coupled receptor) type was found and a phylogenetic tree was constructed with soluble type (sGC) as an outgroup. As in the phylogenetic analysis of Moroz et al, we found a group of GCs with NIT domains that included many metazoans such as placozoa, cnidaria, ecdysozoa, echinoderms and lophotrochozoa, and in Platynereis 15 NIT-GCs were found (Figure 3---figure supplement 1). Using previously published spatially mapped single-cell transcriptome data  [@Achim2015; @williams2017] to find candidate NIT-GCs involved in the cPRC circuit among 15 NIT-GCs, two NIT-GCs (NIT-GC1 and NIT-GC2) were found to be highly expressed in the cPRC (Figure 3---figure supplement 1). Expression analysis with c-opsin1 and MLD pNPs (pedal peptide neuropeptide precursor 2) [@Arendt2004; @williams2017], which have already been identified as cPRC markers using in situ HCR, showed that both NIT-GC1 and NIT-GC2 are strongly and specifically expressed in cPRC (Figure 3A and B, Figure 3---figure supplement 2). To analyse localisations at the protein level, immunostaining was performed using polyclonal antibodies against NIT-GC1 and NIT-GC2. Interestingly, NIT-GC1 was found to localise to the neuronal axon and NIT-GC2 to the ciliary region of the cPRC (Figure 3C and D). Furthermore, co-immunostaining with NOS antibodies revealed that NOS and NIT-GC1 were localised in close proximity in the neurosecretory plexus (Figure 3---figure supplement 3). These results suggest that NIT-GC1 may function as a major target for NO.

@@

NIT-GC1 has a highly conserved cyclase domain and activation of NIT-GC1 is expected to produce cGMP. Recently, a synthetic protein called Green cGull was developed as a cGMP indicator that can be used in cultured cell systems [@Matsuda2017]. We co-expressed Green cGull and NIT-GC1 using the 2A self-cleaving peptide in cultured cells COS-7 (monkey kidney) and tested the intracellular dynamics of cGMP in the activation of NIT-GC1 (Figure 3---figure supplement 4). When the NO donor SNAP was applied to the co-expression cells, the fluorescence intensity in the cells was significantly increased compared to when NIT-GC1 was not expressed (Figure 3E and F). The increase in fluorescence intensity also did not occur when DMSO was applied (Figure 3G). To confirm whether this increase in fluorescence intensity is NIT domain-dependent response of NIT-GC1, mutant form of NIT-GC1 without the NIT domain was expressed and SNAP was applied (Figure 3---figure supplement 4A). As a result, the increase in fluorescence intensity was significantly reduced  (Figure 3H). These results indicate that NIT-GC1 is activated by NO to produce cGMP. They also show that the NIT domain is required for the activation of NIT-GC1.
